---
title: "GadgetR: Gadget and FLR Example"
author: "Ibrahim Umar"
date: "`r Sys.Date()`"
package: gadgetr
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{GadgetR: Gadget and FLR Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# About FLR capability of GadgetR

The FLR project ([https://flr-project.org]) is a collection of fisheries
library in R that provide a set of tools for quantitative fisheries science.

GadgetR has a built-in automatic FLR objects generator feature when running a
model simulation. This interoperability enables users to utilize the wealth of
useful tools that the FLR suite offers when running a Gadget model simulation.

This guide will demonstrate a simple scenario of how to run a simple short-cut
(no assessment) management strategy evaluation (MSE) using a gadget
model as an Operating Model (OM). The management loop, especially the 
recruitment forecast and total allowable catch determination are calculated
using several tools that FLR provides (e.g.,
(FLSR)[https://flr-project.org/doc/Modelling_stock_recruitment_with_FLSR.html]
for the modelling of stock-recruitment). This MSE code also utilizes several
different stock-recruitment models running in parallel to investigate the 
possible uncertainty sources during the simulated management process.

## Installing the package

As for now, the package is yet to find its way into CRAN. However, you can use
our alternative repo that enables you to use R's ```install.packages()```:

```
install.packages("gadgetr", repos = "https://redus-imr.github.io/gadget")
```

Or, you can download and install from the pre-compiled binaries here:

```
https://github.com/REDUS-IMR/gadget/releases/latest
```

Or, you might want to compile it from the source available in Github:

```{r eval=FALSE}
devtools::install_github("REDUS-IMR/gadget", ref="gadgetr")
```

# (Very) simple MSE with GadgetR and FLR

## Setting up the parameters

First we need to define several parameters that will aid the conversion from
gadget output into the FLR objects. This parameters need to be carefully tuned
and reflects the actual gagdet model parametes. 

```{r}
# Set simulation time parameters
firstYear <- 1978
projYear <- 2000
finalYear <- 2020

# Set up a stock (or stocks) for Gadget
## List all the unique species
#stockList <- c("had", "cod")
stockList <- c("had")

# HAD
## List all the fleets for all the species
had.fleets <- c("comm", "survey", "future")

## List all the different stock category for had
had.stocks <- c("had")

## List all the mature stock category for had
had.stocks.mature <- c("had")

## List the survey fleets for had
had.surveys <- c("survey")

## List the forecast fleets for had 
had.forecasts <- c("future")

## Set stock parameters for the OM
# m2 = NULL means we calculate m2 from gadget result, m2 = 0 means we use only
#      residual mortality (m1)
# StockStep: in which step we should observe the stock number
had.params <- list(stockStep = 2, minage = 1, maxage = 10, minfbar = 2, maxfbar = 8,
                   startf = 0.56, endf = 0.65, areas = c(1), m1 = c(0.2), m2 = NULL)

# COD (if applicable)
#cod.fleets <- c("cod.comm", "cod.survey")
#cod.stocks <- c("cod")
#had.params <- list(stockStep = 2, minage = 1, maxage = 8, minfbar = 4, maxfbar = 8,
#                   startf = 0.56, endf = 0.65, areas = c(1), m1 = c(0.2), m2 = NULL)
#...
```

After writing the above (exhaustive) parameters, we need a helper function that can
convert the parameters into a list object that gagdetr libarary can use for
the FLR objects conversion.

```{r}
# Helpers for transforming the above into the new parameter format
parameterize <- function() {
  # Make the global structure
  prm <- list(firstYear = firstYear,
          projYear = projYear,
          finalYear = finalYear,
          stockList = stockList)

  # Process each stock
  for(stock in prm$stockList){
      prm[[stock]] <- list()
      prm[[stock]][["params"]] <- eval(parse(text=paste0(stock, ".params")))
      prm[[stock]][["fleets"]] <- eval(parse(text=paste0(stock, ".fleets")))
      prm[[stock]][["stocks"]] <- eval(parse(text=paste0(stock, ".stocks")))
      prm[[stock]][["stocks.mature"]] <- eval(parse(text=paste0(stock, ".stocks.mature")))
      prm[[stock]][["surveys"]] <- eval(parse(text=paste0(stock, ".surveys")))
      prm[[stock]][["forecasts"]] <- eval(parse(text=paste0(stock, ".forecasts")))
  }
  return(prm)
}
```

## Recruitment prediction function

The following function predicts the next year recruitment given a recruitment
model of choice (e.g., Ricker, Beverton-Holt, etc.) and an FLStock
object. This function utilizes FLR's FLSR tool.

```{r}
# Predict recruitment
predict_recruitment <- function(stk, model) {

	# Create the model
	newFL <- fmle(as.FLSR(stk, model=model))

	# Get the latest ssb
	ssb <- tail(ssb(stk), 1)

	# Predict
  prm <- params(newFL)

	s <- a <- prm[[1]] # Extract 'a' parameter

  if(length(prm) > 1)
	  R0 <- b <- params(newFL)[[2]] # Extract 'b' parameter

  if(length(prm) > 2)
	  v <- c <- params(newFL)[[3]] # Extract 'c' parameter

	# Use the model formula in FLCore
	rec <- eval(parse(text = as.character(model()$model[3])))

	return(list(srmodel = newFL, rec = as.numeric(rec)))
}
```

## Total allowable catch (TAC) prediction function

The following function set the next year TAC given an FLStock object, an
FLIndex object, the year information and an FLR's stock-recruitment model
object. This function utilizes FLR's FLBRP object.

```{r}
# Predict TAC (no assessment, using real data)
predict_TAC <- function(year, stk, idx, srmodel) {

  # Cherry-picked from: https://flr-project.org/doc/An_introduction_to_MSE_using_FLR.html
  # Year needs to be carefully checked

  getCtrl <- function(values, quantity, years, it){
      dnms <- list(iter=1:it, year=years, c("min", "val", "max"))
      arr0 <- array(NA, dimnames=dnms, dim=unlist(lapply(dnms, length)))
      arr0[,,"val"] <- unlist(values)
      arr0 <- aperm(arr0, c(2,3,1))
      ctrl <- fwdControl(data.frame(year=years, quantity=quantity, val=NA))
      ctrl@trgtArray <- arr0
      ctrl
  }

  brp <- brp(FLBRP(stk, srmodel))
  Fmsy <- c(refpts(brp)["msy","harvest"])
  msy <- c(refpts(brp)["msy","yield"])
  Bmsy <- c(refpts(brp)["msy","ssb"])
  Bpa <- 0.5 * Bmsy
  Blim <- Bpa / 1.4

  flag <- ssb(stk)[, as.character(year - 1)] < Bpa
  Ftrgt <- ifelse(flag, ssb(stk)[, as.character(year - 1)] * Fmsy / Bpa, Fmsy)

  sqy <- as.character(c((year - 3):(year-1)))
  fsq <- yearMeans(fbar(stk)[, sqy]) # Use status quo
  ctrl <- getCtrl(c(fsq, Ftrgt), "f", c(year, year + 1), 1)
  stk <- stf(stk, 2)
  gmean_rec <- c(exp(yearMeans(log(rec(stk)))))
  stk <- fwd(stk, control=ctrl, sr=list(model="mean", params = FLPar(gmean_rec,iter=1)))

  TAC <- catch(stk)[, as.character(year)]

	return(TAC)
}
```
## The main MSE loop code

The following code is the main MSE loop code for a single MSE run. This 
function accepts the parameter object and a selected stock-recruitment model
object from the FLCore package (e.g., `bevholt`, `ricker`).

```{r}
runHadSingle <- function(params, modelsel) {

  library(FLCore)
  library(FLBRP)
  library(FLash)
  library(gadgetr)

	# Clean any prior gadget run
	endGadget()

	# Get the path of the example haddock data
	exPath <- loadExample()

	# Init the haddock model
	initGadget(exPath, "refinputfile")

	# Run the hindcast period
	gadgetFLR <- runUntil(params[["projYear"]] - 1, params)

  yrList <- c(params[["projYear"]]:params[["finalYear"]])

	for(year in yrList) {
		print(year)

    # Loop through unique stocks
    for(sp in params$stockList) {
      # Predict and update recruitment
		  pr <- predict_recruitment(gadgetFLR[[sp]]$stk, model = modelsel)
      rec <- round(pr$rec/10000)
		  updateRenewal(sp, year, step = 1, area = 1, age = 1,
            number = rec, mean = 16.41, sdev = 2.25 ,
		        alpha = 8.85e-6, beta = 3.0257)

      # Predict TAC and update catch (spreads throughout 4 seasons/steps)
		  tac <- predict_TAC(year, gadgetFLR[[sp]]$stk, gadgetFLR[[sp]]$idx, pr$srmodel)
      tacPortion <- c(0.232, 0.351, 0.298, 0.119)
      targetFleet <- params[[sp]][["forecasts"]]
      updateAmount(targetFleet, year, 1, 1, tacPortion[[1]] * tac)
      updateAmount(targetFleet, year, 2, 1, tacPortion[[2]] * tac)
      updateAmount(targetFleet, year, 3, 1, tacPortion[[3]] * tac)
      updateAmount(targetFleet, year, 4, 1, tacPortion[[4]] * tac)
    }

    # Forward a year
    out <- runYear()

    # Loop through unique stocks to update the FLR objects
    for(sp in params$stockList) {
      gadgetFLR[[sp]]$stk <- window(gadgetFLR[[sp]]$stk, end = year)
      gadgetFLR[[sp]]$idx <- window(gadgetFLR[[sp]]$idx, end = year)

      gadgetFLR[[sp]] <- updateFLStock(sp, out, as.character(year),
          gadgetFLR[[sp]]$stk, gadgetFLR[[sp]]$idx, params)
    }
	}

  return(gadgetFLR)

}
```

## The driver code for the parallel MSE runs

The main parallel execution code is using `parLapply()` calling
`runHadSingle()` with a list object of different FLCore's
stock-recruitment models and the `params` object.

In the code below we will run the simple MSE with `bevholt`, `ricker`,
`geomean`, `segreg`, `bevholtAR1`, `rickerAR1` and `segregAR1` recruitment
models in parallel. The result of the MSE runs is collated into different
iterations inside an FLStock object.

```{r}
library(parallel)
library(FLCore)
library(ggplotFL)

# Set the number of workers
chk <- Sys.getenv("_R_CHECK_LIMIT_CORES_", "")

if (nzchar(chk) && chk == "TRUE") {
    # use 2 cores in CRAN/actions
    num_workers <- 2L
} else {
    # use all cores in devtools::test() and others
    num_workers <- detectCores()
}

# Make cluster
cl <- makeCluster(num_workers, type = "PSOCK")

# Covert exhaustive list of parameters into a parameter list object
params <- parameterize()

# Export functions and simulation parameter
clusterExport(cl, c("runHadSingle", "predict_TAC", 
                      "predict_recruitment", "params"))

# Try to run with many different recruitment models in parallel
models <- c(bevholt, ricker, geomean, segreg, bevholtAR1,
            rickerAR1, segregAR1)
result <- parLapply(cl, models, function(x) runHadSingle(params, x))

# Stop cluster
stopCluster(cl)

# Make a single FLStock object for the results
had.stk <- result[[1]]$had$stk
had.stk <- propagate(had.stk, length(models))

# Store different run results in different FLStock iterations
for(i in c(1:length(models))) {
  iter(had.stk, i) <- result[[i]]$had$stk
}
```

## Plot the results

Plot the result using FLR's ggplotFL tool.

```{r plot, fig.height=5, fig.width=7, fig.fullwidth=TRUE, fig.cap = "Haddock model prediction using several different stock recruitment models"}
plot(had.stk)
```