---
title: "GadgetR: Quick Start Guide"
author: "Ibrahim Umar"
date: "`r Sys.Date()`"
package: gadgetr
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{GadgetR: Quick Start Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
# Installing the package

As for now, the package is yet to find its way into CRAN. However, you can use
our alternative repo that enables you to use R's ```install.packages()```:

```
install.packages("gadgetr", repos = "https://redus-imr.github.io/gadget")
```

Or, you can download and install from the pre-compiled binaries here:

```
https://github.com/REDUS-IMR/gadget/releases/latest
```

Or, you might want to compile it from the source available in Github:

```{r eval=FALSE}
devtools::install_github("REDUS-IMR/gadget", ref="gadgetr")
```

# Running the example Haddock model

This guide demonstrates gadgetr’s capability of controlling a gadget model run
and extracting relevant information during the run. To run simulations,
GadgetR requires a gadget model as ‘input’ (normally a set of text files). In
this tutorial, we demonstrate how you set up and run GadgetR simulations using
a simple haddock
[model](https://github.com/Hafro/gadget2/blob/master/examples/haddock.md)
by James Begley.

## Loading the package

After installing, the following command will load the library.

```{r setup}
library(gadgetr)
```

## Loading the example model

The library ships with an example haddock model. What we need to do is
calling ```loadExample()``` to obtain the full path of the model.

```{r}
# Get the path of the example haddock data
exPath <- gadgetr::loadExample()
print(exPath)
```

To initialize a gadgetR run, we have to specify the full path of the model
folder and input (parameter) file. In the example model, the input file is
the file named  `refinputfile`.

```{r}
# Initiate the haddock model
gadgetr::initGadget(exPath, "refinputfile")
```

The ```initGadget()``` function instructs gadget to load the model and 
preparing the internal data structure. If successful, the model is now ready
to be run.

## Printing the model ecosystem information

Now that the model is loaded, we can inspect some of the important objects in
the simulation. The ```getEcosystemInfo()``` is a useful function that provides
this type of information.

```{r}
# Get the initial information
initial <- gadgetr::getEcosystemInfo()
print(initial)

# Print start year
print(initial$time["currentYear"])
# Print initial step number
print(initial$time["currentTime"])
# Print final step number
print(initial$time["totalSteps"])
```

## Moving the simulation forward

As we can see in the previous section, the example haddock model has a certain 
number of simulation steps. Now let's loop through the time steps while
collecting the important metrics from the fleets and the stocks.

The package provides two alternative functions to forward the simulation
time. These functions are ```runStep()``` and ```runYear()```, which forward
the simulation by step and year, respectively. Both functions return
the metrics from both stocks and fleets objects at a particular step or year.

```{r}
# Variable for counting step
lengthStep <- 1

# Loop for all steps
stats <- list()
while (TRUE){
  # Append stats
  stats[[lengthStep]] <- gadgetr::runStep()
  # Stop at the end of time
  status <- gadgetr::getEcosystemInfo()
  if(status[["time"]]["finished"] == 1) break
  lengthStep <- lengthStep + 1
}

print(lengthStep)

# Get termination information
term <- gadgetr::getEcosystemInfo()

# Print end year
print(term$time["currentYear"])
print(initial$time["currentYear"] + (lengthStep / 4) - 1)

# Print last step number
print(term$time["currentTime"])
print(lengthStep)
```

In the above code, a condition ```status[["time"]]["finished"] == 1```
terminates the loop. The ```getEcosystemInfo()``` gives an updated status of
the model run whenever it is called. It also gives information whether the
simulation has reached to an end.

You may also safely replace the ```runStep()``` with ```runYear()``` in the
above code. However, the metrics content in the ```stats``` objects will not 
be broke down into steps as you might have with ```runStep()``` function. 
This is because ```runYear()``` function will aggregate the steps' metrics
from the model's objects.

## Inspecting the simulation output

The greatest advantage of this package is its ability to extract and modify
information directly from/to the gadget's internal runtime objects in memory.
This will save users from tinkering with the gadget model configuration files 
and also it output printing files.

This guide won't cover the on-the-fly modification of a gagdet internal
objects. However, the code below should demonstrates the package's capability
to extract gadget information. 

```{r}
# Print out all status
print(names(stats[[1]]))
print(names(stats[[1]]$stock))
print(names(stats[[1]]$fleet))

# Print haddock stock in step 1
print(names(stats[[1]]$stock$had))
print(tail(stats[[1]]$stock$had$stk))

# Print haddock recruitment in step 1
print(names(stats[[1]]$stock$had$rec))
print(tail(stats[[1]]$stock$had$rec$renew))
print(stats[[1]]$stock$had$rec$spawn)

# Print haddock catch by commercial fleet in step 1
print(names(stats[[1]]$fleet))
print(names(stats[[1]]$fleet$comm$catch))
print(tail(stats[[1]]$fleet$comm$catch$had))
```

## Terminating the model run

After each simulation, it is necessary for users to properly terminate an active
gadget simulation run. Use the code below when you finish with a simulation
run or if you want to restart the current model or loading a new model.

```{r}
# Sim cleanup
gadgetr::endGadget()
```

The `endGadget()` function will remove all gadget internal runtime objects from
memory and ensure gadget library is ready to start another simulation run.
