---
title: "GadgetR: (Very) Simple short-cut Management Strategy Evaluation (MSE)
  using Gadget as the Operating Model (OM)"
author: "Ibrahim Umar"
date: "`r Sys.Date()`"
package: gadgetr
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{GadgetR: (Very) Simple short-cut Management Strategy Evaluation (MSE) using Gadget as the Operating Model (OM)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# source: https://stackoverflow.com/questions/23114654
hook_output <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
})
```
This guide will demonstrate how you can use gadgetr to set up and run a simple
short-cut (no assessment) management strategy evaluation (MSE) using a gadget
model as an Operating Model (OM).

# Installing the package

As for now, the package is yet to find its way into CRAN. However, you can use
our alternative repo that enables you to use R's ```install.packages()```:

```
install.packages("gadgetr", repos = "https://redus-imr.github.io/gadget")
```

Or, you can download and install from the pre-compiled binaries here:

```
https://github.com/REDUS-IMR/gadget/releases/latest
```

Or, you might want to compile it from the source available in Github:

```{r eval=FALSE}
devtools::install_github("REDUS-IMR/gadget", ref="gadgetr")
```

# The main code

To facilitate parallel runs, the whole MSE codes are encapsulated in a function
called `runPMSE()`:

```{r, a, eval=FALSE}
runPMSE <- function(seed = 1234) {
```

## Loading the package

Load the package as usual using the command below.

```{r, b, eval=FALSE}
library(gadgetr)
```
## Loading the example model

We will use the example haddock model. For this purpose, we use the model
fitted to the historical period (1978–1999) and project 21-year forecasts
(2000–2020) by extending the simulation years in the gadget model.This means
that the model end time parameter (`lastyear`) is set to 2020.

```{r, c, eval=FALSE}
# Get the path of the example haddock model
exPath <- gadgetr::loadExample()
print(exPath)

# Initiate the haddock model
gadgetr::initGadget(exPath, "refinputfile")

# Print the basic model information
gadgetr::getEcosystemInfo()

# Init seed
set.seed(seed)
```

## Defining helper functions

For this simple demonstration, we create four helper functions for MSE
simulations; 1) a function to calculate SSB, 2) a function to extract recruit
numbers, 3) a function to forecast recruitment, and 4) a function to calculate
TAC.

### 1) Calculating SSB

This function calculates the spawning stock biomass (SSB) for a given stock.
Here we assume that individuals of 3-year-olds and older are mature and only
stock information in step 4 will be taken into the calculation.

```{r, d, eval=FALSE}
# Calculate SSB
calcSSB <- function(stk) {
    # Get SSB
    subs <- stk[stk[, "age"] >= 3 & stk[, "step"] == 4, ]
    ssb <- sum(subs[, "number"] * subs[, "meanWeights"])
    return(ssb)
}
```

### 2) Calculating recruitment

A simple function for calculating the "real" recruitment values (no observation
error) for a given stock in a year. Only 1-year-olds are used in the
calculation. Because stock information (numbers and weights) is collected after
each quarter (`step`) in simulations, we use the recruitment values from
quarter 2.

```{r, e, eval=FALSE}
# Calculate Recruitment
calcRecruitment <- function(stk) {
    # Get Recruitment (based on fbar and only from the last step)
    subs <- stk[stk[, "age"] == 1 & stk[, "step"] == 2, ]
    return(sum(subs[, "number"]))
}
```
### 3) Forecasting a recruitment

This function uses the constant recruitment model fit to forecast recruitment.

```{r, f, eval=FALSE}
# In this example, we assume that the haddock recruitment for a given year is
# equal to the historical geometric mean recruitment
fcRecruitment <- function(rec) {
    rec <- unlist(rec)
    return(exp(mean(log(rec))))
}
```

### 4) Calculating TAC

We use a simple function to project the next year’s total allowable catch
(TAC), assuming that target exploitation rate is 40% of stock biomass in a
given year.

```{r, g, eval=FALSE}
# and calculating TAC is a matter of taking 40% of the stock's last year
# stock biomass value with added noise
calcTAC <- function(stk) {
    # Get TSB
    subs <- stk[stk[, "step"] == 4, ]
    tsb <- sum(subs[, "number"] * subs[, "meanWeights"])
    # Calculate error
    err <- rnorm(1, 0, tsb * .25)
    # Calculate TAC
    tac <- .4 * (tsb + err)
    return(tac)
}
```

## Running the hindcast simulation

The haddock model’s hindcast period spans from 1978 to 1999. As the first step,
we run gadget until 1999.

```{r, h, eval=FALSE}
# Placeholders for values
stats <- list()
ssb <- list()
rec <- list()

# Loop for the hindcast period
forecastYear <- 2000
status <- gadgetr::getEcosystemInfo()
while ( status[["time"]]["currentYear"] < forecastYear ) {
  # Append stats
  curYr <- as.character(status[["time"]]["currentYear"])
  stats[[curYr]] <- gadgetr::runYear()
  ssb[[curYr]] <- calcSSB(stats[[curYr]]$stock$had$stk)
  rec[[curYr]] <- calcRecruitment(stats[[curYr]]$stock$had$stk)
  # Get the latest runtime information
  status <- gadgetr::getEcosystemInfo()
}
```

## Running the forecast simulation

Now, we run the simulation from year 2000 to 2020. In looping through the
years, we also perform the following steps at the beginning of the year
(i.e., before quarter (`step`) 1):

1. Project recruitment by calling `fcRecruitment()`.
2. Apply the projected recruitment to the simulation using
`gadgetr::updateRenewal()` on the _had_ stock.
3. Calculate the last year’s SSB at the beginning of the year using
`calcSSB()`.
4. Calculate TAC for the year using `calcTAC()`.
5. Apply the calculated TAC to the _future_ total fleet in the gadget
simulation using `gadgetr::updateAmount()`.

```{r, i, eval=FALSE}
# MSE loop
while ( status[["time"]]["finished"] != 1 ) {
  # Time calculation
  curYr <- status[["time"]]["currentYear"]
  lastYr <- curYr - 1
  
  # Calculate Recruitment for this year
  recNumber <- fcRecruitment(rec)
  rec[[as.character(curYr)]] <- recNumber
  print(paste("Recruitment in", curYr, "is", recNumber))

  # Apply recruitment number for this year haddock stock
  # note that this is similar to the format found in "had.rec" file
  updateRenewal("had", curYr, step = 1, area = 1, age = 1,
      number = (recNumber/10000), mean = 16.41, sdev = 2.25 , alpha = 8.85e-6,
      beta = 3.0257)

  # Calculate last year SSB
  stk0 <- stats[[as.character(lastYr)]]$stock$had$stk
  ssb[[as.character(lastYr)]] <- calcSSB(stk0)

  # Calculate TAC for this year
  tac <- calcTAC(stk0)
  print(paste("TAC in", curYr, "is", tac))

  # Apply TAC to fleet data (spread out over 4 quarter/step in a year)
  # in percentage (i.e., sum(tacPortion) == 1)
  tacPortion <- c(0.232, 0.351, 0.298, 0.119)
  targetFleet <- "future"
  updateAmount(targetFleet, curYr, 1, 1, tacPortion[[1]] * tac)
  updateAmount(targetFleet, curYr, 2, 1, tacPortion[[2]] * tac)
  updateAmount(targetFleet, curYr, 3, 1, tacPortion[[3]] * tac)
  updateAmount(targetFleet, curYr, 4, 1, tacPortion[[4]] * tac)

  # Forward the time
  stats[[as.character(curYr)]] <- gadgetr::runYear()

  # Get the latest runtime information
  status <- gadgetr::getEcosystemInfo()
}
```

## Summarizing stock status and catch

After the loop finishes, we summarize the main performance metrics.

```{r, j, eval=FALSE}
print(rec)
# Calculate real catch (all fleets)
catch <- lapply(stats, function(x) {
    sum(x$fleet$future$catch$had[, "biomassConsumed"], na.rm = TRUE) +
    sum(x$fleet$comm$catch$had[, "biomassConsumed"], na.rm = TRUE) +
    sum(x$fleet$survey$catch$had[, "biomassConsumed"], na.rm = TRUE)
  })
```

## Ending gadget

After each simulation, it is necessary to properly terminate an active
gadget simulation run.

```{r, k, eval=FALSE}
endGadget()
```

## Ending runPMSE

```{r, l, eval=FALSE}
    return(list(ssb, rec, catch))
}
```

This is the end of the `runPMSE()` and we return the recruitment, SSB and 
catch metrics.

# Do a single run

Now let's try to do a single run and plot the output. But first let's make a
plot generator function.

```{r}
plotIt <- function(x) {
    singlePlot <- function(cat, name, title) {
        # Collect metrics
        z <- do.call(cbind, lapply(x, function(y) unlist(y[[cat]])))
        tab <- as.data.frame(z)
        tab$year <- as.numeric(rownames(tab))
        tab <- reshape(tab, varying=list(1:(ncol(tab)-1)), direction="long", v.names = name)
        fig <- ggplot(tab, aes_string("year", name)) +
                    stat_summary(geom = "line", fun = mean) +
                    stat_summary(geom = "ribbon", fun.data = mean_sdl, fun.args = list(mult = 1), fill = "red", alpha = 0.2) +
                    stat_summary(geom = "ribbon", fun.data = mean_cl_normal, fill = "red", alpha = 0.3) + 
                    labs(title=paste("Haddock", title), y="")
        return(fig)
    }

  ssbplot <- singlePlot(1, "ssb", "SSB")
  recplot <- singlePlot(2, "rec", "Recruitment")
  catchplot <- singlePlot(3, "catch", "Catch")

  return(list(ssbplot, recplot, catchplot))
}
```

```{r ref.label=c('a','b','c','d','e','f','g','h','i','j','k','l'), echo=FALSE}
```
```{r output.lines=25}
library(ggplot2)
result <- runPMSE()
```

## Plotting the result

Generate the performance metric plots.

```{r plot1, fig.height=5, fig.width=7, fig.fullwidth=TRUE, fig.cap="Haddock model (single-run) performance metrics"}
plots <- plotIt(list(result))
print(plots[[1]])
print(plots[[2]])
print(plots[[3]])
```

# Do multiple runs in parallel

Here is an example of doing a multiple iteration runs utilizing the R parallel
execution capability.

```{r}
library(parallel)

chk <- Sys.getenv("_R_CHECK_LIMIT_CORES_", "")

if (nzchar(chk) && chk == "TRUE") {
    # use 2 cores in CRAN/actions
    num_workers <- 2L
} else {
    # use all cores in devtools::test()
    num_workers <- detectCores()
}

cl <- makeCluster(num_workers, type = "PSOCK")

# Run
seeds <- round(runif(100, 1, 10000))
result <- parLapply(cl, seeds, runPMSE)

stopCluster(cl)

plots <- plotIt(result)

```

## Plotting the result

Plot the performance metrics from the runs.

```{r plot2, fig.height=5, fig.width=7, fig.fullwidth=TRUE, fig.cap="Haddock model (parallel-runs) performance metrics"}
print(plots[[1]])
print(plots[[2]])
print(plots[[3]])
```




