---
title: "GadgetR: Quick and Dirty Management Strategy Evaluation (MSE) using
  Gadget as the Operating Model (OM)"
author: "Ibrahim Umar"
date: "`r Sys.Date()`"
package: gadgetr
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{GadgetR: Quick and Dirty Management Strategy Evaluation (MSE) using Gadget as the Operating Model (OM)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# source: https://stackoverflow.com/questions/23114654
hook_output <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
})
```
This guide will demonstrate the gadgetr ability to run and manipulate a gadget
model simulation so that it would perform as an Operating Model (OM) in a
Management Strategy Evaluation (MSE) loop.

# Loading the package

Load the package as usual using the command below.

```{r setup}
library(gadgetr)
```
# Loading the example model

For this guide we will use the included haddock model. This model contains the
observed stock and fleet data up to year 1999. However, the model end time
(i.e., `lastyear` parameter) has been set as 2020 to support an exploratory
forecast run.

```{r}
# Get the path of the example haddock data
exPath <- gadgetr::loadExample()
print(exPath)

# Init the haddock model
gadgetr::initGadget(exPath, "refinputfile")

# Print info
gadgetr::getEcosystemInfo()

# Init seed
set.seed(1098765)
```

# Defining helper functions

For the MSE purpose, we need to define four helper functions, namely a SSB
calculation function, a recruitment calculation function, a recruitment
forecast function, and a TAC calculation function.

## Calculating SSB

This function calculates the spawning stock biomass (SSB) for a given stock.
Here we assume that individuals with age more than 3 are mature and only stock
information in step 4 will be taken into the calculation.

```{r}
# Calculate SSB
calcSSB <- function(stk) {
    # Get SSB
    subs <- stk[stk[, "age"] >= 3 & stk[, "step"] == 4, ]
    ssb <- sum(subs[, "number"] * subs[, "meanWeights"])
    return(ssb)
}
```

## Calculating recruitment

A simple function for calculating the "real" recruitment values for a given
stock in a year. Only age 1 individuals are used in the calculation
and as stock information (numbers and weights) were collected after each step
is run, we use only the values from step 2.

```{r}
# Calculate Recruitment
calcRecruitment <- function(stk) {
    # Get Recruitment (based on fbar and only from the last step)
    subs <- stk[stk[, "age"] == 1 & stk[, "step"] == 2, ]
    return(sum(subs[, "number"]))
}
```
## Forecasting a recruitment

This function use a bit of "magic" to forcasting the stok recruitment value.

```{r}
# Supposed the haddock recruitment forecast for a given year is a matter of
# picking a random number...
fcRecruitment <- function() {
    return(sample(1:25, 1) * 1000)
}
```

## Calculating TAC

A simple function for calculating the next year's total allowable catch (TAC)
given a stock's SSB value for a particular year.

```{r}
# ... and calculating TAC is a matter of taking 35% of the stock's last year
# SSB value
calcTAC <- function(ssb) {
    return(.4 * ssb)
}
```

# Running the hindcast simulation

The haddock example model hindcast period spans from year 1978 until 1999.
Therefore, as a first step, we need to run gadget until the end of year 1999.

```{r}
# Placeholders for values
stats <- list()
ssb <- list()
fcrec <- list()
rec <- list()

# Loop for the hindcast period
forecastYear <- 2000
status <- gadgetr::getEcosystemInfo()
while ( status[["time"]]["currentYear"] < forecastYear ) {
  # Append stats
  curYr <- as.character(status[["time"]]["currentYear"])
  stats[[curYr]] <- gadgetr::runYear()
  ssb[[curYr]] <- calcSSB(stats[[curYr]]$stock$had$stk)
  # Get the latest runtime information
  status <- gadgetr::getEcosystemInfo()
}
```

# Running the forecast simulation

Now, we run the simulation from year 2000 until 2020. In addition to looping
through the years, we also performing these additional steps at the beginning
of the year (i.e., before step 1):

1. Defining the Haddock recruitment value by calling the `fcRecruitment()`
function.
2. Applying the obtained recruitment value into gadget simulation using
`gadgetr::updateRenewal()` function, targeting the "had" stock.
3. Calculating last year's SSB value in the beginning of the year using
`calcSSB()` function.
4. Calculating TAC value for the year using `calcTAC()` function.
5. Applying the calculated TAC value back into gadget simulation using
`gadgetr::updateAmount()` function, targeting the "future" fleet.

```{r output.lines=25}
# MSE loop
while ( status[["time"]]["finished"] != 1 ) {
  # Time calculation
  curYr <- status[["time"]]["currentYear"]
  lastYr <- curYr - 1
  
  # Calculate Recruitment for this year
  recNumber <- fcRecruitment()
  fcrec[[as.character(curYr)]] <- recNumber
  print(paste("Recruitment in", curYr, "is", recNumber))

  # Apply recruitment number for this year haddock stock
  # note that this is similar to the format found in "had.rec" file
  updateRenewal("had", curYr, step = 1, area = 1, age = 1,
      number = recNumber, mean = 16.41, sdev = 2.25 , alpha = 8.85e-6,
      beta = 3.0257)

  # Calculate last year SSB
  ssb[[as.character(lastYr)]] <- 
      calcSSB(stats[[as.character(lastYr)]]$stock$had$stk)

  # Calculate TAC for this year
  tac <- calcTAC(ssb[[as.character(lastYr)]])
  print(paste("TAC in", curYr, "is", tac))

  # Apply TAC to fleet data (spread out between 4 steps in a year)
  tacPortion <- c(0.232, 0.351, 0.298, 0.119)
  targetFleet <- "future"
  updateAmount(targetFleet, curYr, 1, 1, tacPortion[[1]] * tac)
  updateAmount(targetFleet, curYr, 2, 1, tacPortion[[2]] * tac)
  updateAmount(targetFleet, curYr, 3, 1, tacPortion[[3]] * tac)
  updateAmount(targetFleet, curYr, 4, 1, tacPortion[[4]] * tac)

  # Forward the time
  stats[[as.character(curYr)]] <- gadgetr::runYear()

  # Get the latest runtime information
  status <- gadgetr::getEcosystemInfo()
}
```

# Summarizing recruitment and catch values

After the loop finishes, we might want to summarize several important metric
values over the years.

```{r}
# Calculate real recruitment
rec <- lapply(stats, function(x) calcRecruitment(x$stock$had$stk))

# Calculate real catch (future fleet)
catch <- lapply(stats, function(x) 
    sum(x$fleet$future$catch$had[, "biomassConsumed"]))

```

# Plotting the results

And plot it:

```{r plot, fig.height=5, fig.width=7, fig.fullwidth=TRUE, fig.cap = "Haddock simulation numbers"}
# Plot the world
maxy <- max(c(unlist(ssb), unlist(rec)))
cl <- rainbow(3)
plot(0, 0, xlim = c(1978, 2020), ylim = c(0, maxy), type = "n", xlab = "Year",
    ylab = "", main = "Haddock gadget model")
lines(names(ssb), ssb, col = cl[1], type = "l")
lines(names(rec), rec, col = cl[2], type = "l")
lines(names(catch), catch, col = cl[3], type = "l")
abline(v = 2000, col="darkorange", lwd=3, lty=2)
legend(1980, maxy, legend=c("SSB (had)", "Recruitment (had)", 
    "Catch (future fleet)"), col=cl, lty=1, cex=1, bg='lightblue1')
```

# Ending gadget

After each simulation, it is necessary to properly terminate an active
gadget simulation run.

```{r}
endGadget()
```


